---
title: "MILESTONE EXAM 1 - STUDENT TEMPLATE"
author: "Shirisha Biyyala"
date: "Due: October 2, 2024"
output: 
  github_document: 
    fig_height: 4
    fig_width: 7
---

```{r setup, include = FALSE}
# set the echo option to FALSE to see how the document looks with the code suppressed
knitr::opts_chunk$set(echo = TRUE)
```

## Academic Honesty Statement

I, Shirisha Biyyala, hereby state that I have not communicated with or gained information in any way from my classmates or anyone other than the Professor or TA during this exam, and that all work is my own.

## Load packages

```{r load-packages, message=FALSE}
# load required packages here
library(tidyverse)
library(nycflights13)
library(ggplot2)
library(dplyr)
library(knitr)
library(htmltools)
```

## Questions 

This section allows us to view the entire `flights` dataset from the `nycflights13` package. It also provides access to documentation for understanding the datasets' variables and attributes.


```{r understand-data}
#view(flights)
#?flights
```

In this section, we extract the first few rows of the `flights` dataset to get a **glimpse of the data**. 

```{r flights-head}
# Get the head of the dataset
flights_head <- head(flights)

# Create an HTML table
html_table <- knitr::kable(flights_head, format = "html", escape = FALSE)

# Print the scrollable table
HTML(paste0(
  "<div style='overflow-x: auto; max-width: 100%; height: 300px;'>",
  html_table,
  "</div>"
))
```

This section conducts an initial **exploratory analysis** by checking the structure of the `flights` dataset using glimpse, which provides a compact view of the data types and sample values. It counts the number of missing values in each column to identify potential issues with data completeness. Finally, **summary statistics** are generated for key features like departure delay, arrival delay, distance, and airtime, allowing for a better understanding of flight performance and trends.


```{r exploratory-data-analysis}
# Check structure and missing values
glimpse(flights)

# Count missing values in each column
sapply(flights, function(x) sum(is.na(x)))

# Summary statistics for key features
summary(flights %>% select(dep_delay, arr_delay, distance, air_time))
```

### Question 1

```{r most-common-destinations}
```

### Question 2

[Enter code and narrative here.]

### Question 3

[Enter code and narrative here.]

### Question 4

[Enter code and narrative here.]

### Question 5

[Enter code and narrative here.]

### Question 6

[Enter code and narrative here.]

### Question 7

[Enter code and narrative here.]

### Question 8

[Enter code and narrative here.]

### Extra Credit

[Enter code and narrative here.]


