---
title: "MILESTONE EXAM 1 - STUDENT TEMPLATE"
author: "Shirisha Biyyala"
date: "Due: October 2, 2024"
output: 
  github_document: 
    fig_height: 4
    fig_width: 7
---

```{r setup, include = FALSE}
# set the echo option to FALSE to see how the document looks with the code suppressed
knitr::opts_chunk$set(echo = TRUE)
```

## Academic Honesty Statement

I, Shirisha Biyyala, hereby state that I have not communicated with or gained information in any way from my classmates or anyone other than the Professor or TA during this exam, and that all work is my own.

## Load packages

```{r load-packages, message=FALSE}
# load required packages here
library(tidyverse)
library(nycflights13)
library(ggplot2)
library(dplyr)
library(knitr)
library(htmltools)
```

## Questions 

#### Understand the dataset

This section allows us to view the entire `flights` dataset from the `nycflights13` package. It also provides access to documentation for understanding the datasets' variables and attributes.


```{r understand-data}
#view(flights)
#?flights
```

In this section, we extract the first few rows of the `flights` dataset to get a **glimpse of the data**. 

```{r flights-head}
# Get the head of the dataset
flights_head <- head(flights)

# Create an HTML table
html_table <- knitr::kable(flights_head, format = "html", escape = FALSE)

# Print the scrollable table
HTML(paste0(
  "<div style='overflow-x: auto; max-width: 100%; height: 300px;'>",
  html_table,
  "</div>"
))
```

#### Exploratory Data Analysis

This section conducts an initial **exploratory analysis** by checking the structure of the `flights` dataset using glimpse, which provides a compact view of the data types and sample values. It counts the number of missing values in each column to identify potential issues with data completeness. Finally, **summary statistics** are generated for key features like departure delay, arrival delay, distance, and airtime, allowing for a better understanding of flight performance and trends.


```{r exploratory-data-analysis}
# Check structure and missing values
glimpse(flights)

# Count missing values in each column
sapply(flights, function(x) sum(is.na(x)))

# Summary statistics for key features
summary(flights %>% select(dep_delay, arr_delay, distance, air_time))
```

### Question 1

To identify the **ten most common flight destinations** from NYC airports in 2013, I grouped the flights dataset by the `dest` column and counted the number of flights to each destination. After *sorting* the results in descending order, I selected the top ten destinations.


```{r most-common-destinations}
flights %>% 
  group_by(dest) %>% 
  summarise(flight_count = n()) %>% 
  arrange(desc(flight_count)) %>% 
  top_n(10)   # Select the top 10 destinations
```

From the table, we can see that **ORD** had the highest number of flights from NYC, followed closely by **ATL** and **LAX**. This indicates that these cities were the most popular destinations for travelers flying out of New York City in 2013.

### Question 2

To determine which airlines had the **most flights departing** from NYC airports in 2013, I grouped the `flights` dataset by the `carrier` column and calculated the total number of flights for each airline. I then joined this data with the airlines dataset to retrieve the airline names, sorted the results in descending order, and presented the flight counts alongside their respective airline names.


```{r most-flights-departing}
flights %>% 
  group_by(carrier) %>% 
  summarise(flight_count = n()) %>% 
  arrange(desc(flight_count)) %>%            
  left_join(airlines, by = "carrier") %>% 
  rename(airline_name = name)
```

From the table, we can see that **United Air Lines Inc.** (58665) had the highest number of flights from NYC, followed by **JetBlue Airways**(54635) and **ExpressJet Airlines Inc.**(54173) This highlights the dominance of these airlines in the NYC air travel market during 2013.

### Question 3

To analyze the **arrival delays of flights** with non-missing delay information, I filtered the `flights` dataset to exclude any rows where the `arr_delay` value is missing. I then grouped the data by the `carrier` column, calculating the *mean arrival delay for each airline*. After sorting the results, I identified the carrier with the **highest and lowest mean arrival delays**, along with their corresponding carrier codes.

```{r arrival-delays}
flights %>% 
  filter(!is.na(arr_delay)) %>% 
  group_by(carrier) %>% 
  summarise(mean_arr_delay = round(mean(arr_delay), 4)) %>%
  left_join(airlines, by= "carrier") %>% 
  arrange(desc(mean_arr_delay))
```

#### Highest mean arrival delay
The carrier with the highest mean arrival delay is **Frontier Airlines Inc. (F9)**, with an average delay of **21.92 minutes**. This indicates that, on average, flights operated by Frontier Airlines arrived nearly **22 minutes later** than their scheduled time, suggesting potential operational challenges affecting punctuality.


#### Lowest mean arrival delay
Conversely, **Alaska Airlines Inc. (AS)** had the lowest mean arrival delay, averaging **-9.93 minutes**. This negative value indicates that, on average, Alaska Airlines flights arrived about **10 minutes earlier** than scheduled. This suggests that Alaska Airlines was the most punctual among the carriers analyzed, performing well in terms of timely arrivals.


### Question 4

[Enter code and narrative here.]

### Question 5

[Enter code and narrative here.]

### Question 6

[Enter code and narrative here.]

### Question 7

[Enter code and narrative here.]

### Question 8

[Enter code and narrative here.]

### Extra Credit

[Enter code and narrative here.]


