---
title: "HW 05 - Legos"
author: "Shirisha Biyyala"
date: " `r Sys.Date()` "
output: github_document
---

## Load packages and data

```{r load-packages-and-data, message = FALSE}
#install.packages("tidyverse")
library(tidyverse)
load("data/lego_sales.rda")
```

```{r view-data}
view(lego_sales)
glimpse(lego_sales)

# Get the head of the data set
lego_head <- head(lego_sales)
lego_head

# Get the tail of the data set
lego_tail <- tail(lego_sales)
lego_tail
```


## Exercises

### Exercise 1

In this sample, the first three common names of purchasers are **Jackson**, **Jacob** and **Joseph**.

```{r common-first-names}
lego_sales %>%
  count(first_name) %>%
  arrange(desc(n)) %>%
  slice_max(n, n = 3)
```


### Exercise 2

Remove this text, and add your answer for Exercise 2 here.
Add code chunks as needed.
Don't forget to label your code chunk.
Do not use spaces in code chunk labels.


### Exercise 3

Remove this text, and add your answer for Exercise 3 here.
Add code chunks as needed.
Don't forget to label your code chunk.
Do not use spaces in code chunk labels.

### Exercise 4

...

### Exercise 5

...

### Exercise 6

...

### Exercise 7

...

### Exercise 8

...

### Exercise 9

...
