---
title: "HW 01 - Pet names"
author: "Shirisha Biyyala"
date: "`r Sys.Date()`"
output: github_document
---

## Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse)
library(openintro)

```

## Exercises

### Exercise 1

The data dictionary indicates that the dataset contains 52,519 pets.

```{r count_rows}
total_rows <- nrow(seattlepets)
total_rows
```


### Exercise 2

According to the data dictionary, each pet is described by 7 variables.

```{r count_cols}
total_cols <- ncol(seattlepets)
total_cols
```


### Exercise 3

The three most popular pet names are Lucy, with a frequency of 439; Charlie, with 387; and Luna, with 355. Additionally, there are 483 pet names listed as unknown, represented as NA.

```{r most-common-names}
seattlepets %>%
  count(animal_name, sort = TRUE)
```

###     Exercise 4

By updating the code, we identified the most common pet names for each species. We used slice_head() to find the top names based on their frequency.

Results:

1. Cats: The top names are Lucy (111 occurrences), Luna (102), Lily (86), and Max (83).
2. Dogs: The most common names are Lucy (337), Charlie (306), Bella (249), Luna (244), and Daisy (221).
3. Goats: The name Heidi (2 occurrences) is the most common, showing generally low frequencies for goat names.
4. Pigs: Each pig name appears only once, so all names are listed with the same frequency.

```{r most-common-cat-dog-names}
seattlepets %>%
group_by(species) %>%
count(animal_name, sort = TRUE) %>%
slice_head(n = 5)
```

### Exercise 5

Remove this text, and add your answer for Exercise 5 here.

### Exercise 6

Remove this text, and add your answer for Exercise 5 here.
