---
title: "HW 04 - What should I major in?"
author: "Shirisha Biyyala"
date: "2024-09-30"
output: github_document
---

## Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse)
library(scales)
library(fivethirtyeight)
```

```{r understand-data}
#glimpse(college_recent_grads)
#?college_recent_grads
#view(college_recent_grads)
```

```{r college_recent_grads_head, echo=FALSE, results='asis'}
# Get the head of the dataset
college_recent_grads_head <- head(college_recent_grads)

# Create an HTML table
html_table <- knitr::kable(college_recent_grads_head, format = "html", escape = FALSE)

# Print the scrollable table
cat(paste0(
  "<div style='overflow-x: auto; max-width: 100%; height: 300px;'>",
  html_table,
  "</div>"
))
```

## Exercises

### Exercise 1

In this code chunk, we will:

- Arrange the data by the proportion of women (sharewomen) in descending order,highlighting majors with the highest proportion of women at the top.
- Select only the relevant columns: major, total (the number of people with that major),and sharewomen (the proportion of women).
- Display only the top 3 majors by using top_n(3).

```{r top_n_major}
college_recent_grads %>%
  arrange(desc(sharewomen)) %>% 
  select(major, total, sharewomen) %>%
  top_n(3)
```


### Exercise 2

In this code chunk, we will:

- **Visualize Median Income Distribution:** Create a histogram to illustrate how median incomes are distributed among recent graduates by major.
  
- **Select Binwidth:** Use a binwidth of $1000 to capture meaningful variations in income without excessive detail.

- **Analyze Distribution Characteristics:** Examine the histogram to identify patterns, skewness, and potential outliers in the income data.

```{r histogram-visualization}
ggplot(data = college_recent_grads, mapping = aes(x = median)) +
  geom_histogram() +
  theme_minimal()


# Create histogram with a binwidth of $1000
ggplot(data = college_recent_grads, mapping = aes(x = median)) +
  geom_histogram(binwidth = 1000, fill = "lightblue") +
  labs(title = "Distribution of Median Incomes (Binwidth = $1000)",
       x = "Median Income",
       y = "Count") +
  theme_minimal()

# Create histogram with a binwidth of $5000
ggplot(data = college_recent_grads, mapping = aes(x = median)) +
  geom_histogram(binwidth = 5000, fill = "rosybrown") +
  labs(title = "Distribution of Median Incomes (Binwidth = $5000)",
       x = "Median Income",
       y = "Count") +
  theme_minimal()
```

### Exercise 3

Remove this text, and add your answer for Exercise 3 here.
Add code chunks as needed.
Don't forget to label your code chunk.
Do not use spaces in code chunk labels.

### Exercise 4

...

### Exercise 5

...

### Exercise 6

...

### Exercise 7

...

### Exercise 8

...

### Exercise 9

...

### Exercise 10

...
