---
title: "HW 03 - Road traffic accidents"
author: "Shirisha Biyyala"
date: "2024-09-25"
output: github_document
---

## Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse)

load("data/accidents.rda")
```

## Exercises

### Exercise 1

The dataset has `r nrow(accidents)` observations.

### Exercise 2

Each row in the dataset represents an individual accident or crash incident, with each column providing specific details about that event. These details typically include:

- **Date/Time** of the accident
- **Location** (city, intersection, road)
- **Accident type** (collision type)
- **Vehicles involved**
- **Weather/Road conditions**
- **Casualties** (injuries, fatalities, damages)
- **Other contributing factors** (e.g., speeding, alcohol)

It's a record of specific accident events with all relevant information.

```{r}
view(accidents)
```

```{r, echo=FALSE, results='asis'}
# Get the head of the dataset
accidents_head <- head(accidents)

# Create an HTML table
html_table <- knitr::kable(accidents_head, format = "html", escape = FALSE)

# Print the scrollable table
cat(paste0(
  "<div style='overflow-x: auto; max-width: 100%; height: 300px;'>",
  html_table,
  "</div>"
))
```


### Exercise 3

Remove this text, and add your answer for Exercise 3 here.

```{r label-me-1, fig.fullwidth = TRUE}
accidents %>%
  mutate(day_of_week_type = if_else(day_of_week %in% c("Saturday", "Sunday"), "Weekend", "Weekday"))
# remove this comment and add the code for Exercise 3 here
```

### Exercise 4

Remove this text, and add your answer for Exercise 4 here.

```{r label-me-2}
# remove this comment and add the code for Exercise 4 here
```
