---
title: "hw08 - Maps"
author: "Shirisha Biyyala"
date: "2024-11-27"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r load-packages, echo=TRUE, message = FALSE, warning = FALSE}
library(tidycensus)
library(dplyr)
library(tigris)
library(ggplot2)
library(sf)
```


## Set up and verify API Key 

```{r api-key}
# census_api_key("YOUR_API_KEY_removed", install = TRUE)

test_data <- get_acs(geography = "state", variables = "B01003_001", year = 2022)
head(test_data)
```

API key has been setup and the data preview indicates that the API is working correctly.

## Load variables 

```{r code-variables}
# Load the list of variables 
variables <- load_variables(year = 2022, dataset = "acs1")

# Preview the variables
head(variables)
```

### Part 1: State-Level Data and Mapping

#### Step 1: Load State-Level Data

```{r download-data}
# Download state-level population data
state_data <- get_acs(geography = "state", variables = "B01003_001", year = 2022, survey = "acs1")

# Preview the data
head(state_data)
```

#### Step 2: Download US State Shapefiles

```{r shape-files}
# Download state shapefiles
state_shapes <- states(cb = TRUE)
```

#### Step 3: Merge Data Using State Names

```{r merge-data}
# Convert state names to lowercase for consistency
state_data <- state_data %>%
  mutate(NAME = tolower(NAME))
state_shapes <- state_shapes %>%
  mutate(NAME = tolower(NAME))

# Merge the data using state names
state_map_data <- left_join(state_shapes, state_data, by = "NAME")

# Check the merged data
head(state_map_data)
```

#### Step 4: Plot the US Map

```{r explore-data}
glimpse(state_map_data)
```

##### Step 4.1: Your first map

```{r basic-map}
# Create the map
state_map_data %>%
  ggplot(aes(fill = estimate)) +
    geom_sf()  +
    coord_sf(crs = st_crs(5070))  +
    labs(title = "Population by State (2022)", fill = "Population") +
    theme_minimal()
```

##### Step 4.2: Filter Data by Geographical Bounds

```{r contiguous_boundaries}
# Filter for contiguous states using latitude and longitude bounds
states_clean <- state_map_data %>%
  filter(!NAME %in% c("alaska", "american samoa", "hawaii", "guam", "puerto rico", 
                      "commonwealth of the northern mariana islands", 
                      "united states virgin islands"))
```

##### Step 4.3: Change projection

```{r change-projection}
# Change projection to Albers Equal Area
state_map <- ggplot(states_clean) +
  geom_sf(aes(fill = estimate)) +
  coord_sf(crs = st_crs(5070)) +
  labs(title = "Population by Contiguous U.S. States (2022)", fill = "Population") +
  theme_minimal()

state_map
```

##### Step 4.4: Remove axes and grid

```{r clean-visual}
# Final map without axes and gridlines
ggplot(states_clean) +
  geom_sf(aes(fill = estimate)) +
  labs(title = "Population by State (2022)", fill = "Population") +
  coord_sf(crs = st_crs(5070)) +  # Apply projection
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank()
  )
```

### Part 2: County-Level Data and Mapping for a Specific State

#### Step 1: Load County-Level Data

```{r county-data}
# Create the map
# Download county-level population data for California
county_data <- get_acs(geography = "county", variables = "B01003_001", state = "CA", year = 2022)

# Preview the data
head(county_data)
```

#### Step 2: Download County Shapefiles

```{r cal-county-shape}
# Download county shape files for California
county_shapes <- counties(state = "CA", cb = TRUE)
head(county_shapes)
```

```{r check-names}
# Check the names in both datasets
head(county_data$NAME)
head(county_shapes$NAME)
```

```{r clean-county-shape}
# Clean the 'NAME' column in county_shapes by removing ' County, California' and convert to lowercase
county_shapes$NAME <- tolower(gsub(" County, California", "", county_shapes$NAME))

# Check if cleaning has worked
head(county_shapes$NAME)
```
```{r clean-county-data}
# Clean the 'NAME' column in county_data by removing ' county, california' and convert to lowercase
county_data$NAME <- tolower(gsub(" county, california", "", county_data$NAME))

# Check if cleaning has worked
head(county_data$NAME)
```



#### Step 3: Merge Data Using County Names

```{r merge-county-data}
# Merge the data by county name
county_map_data <- left_join(county_shapes, county_data, by = "NAME")

# Check the merged data
head(county_map_data)
```

#### Step 4: Plot the County Map

```{r county-map}
# Create the map with county-level population data
ggplot(county_map_data) +
  geom_sf(aes(fill = estimate)) +
  labs(title = "Population by County in California (2022)", fill = "Population") +
  coord_sf(crs = st_crs(5070)) +  # Apply projection
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank()
  )
```





