---
title: "hw09"
author: "Shirisha Biyyala"
date: "2024-12-04"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Linear Models and Fixed Effects in R

### Prerequisites

```{r load-packages, message = FALSE, warning = FALSE}
library(gapminder)
library(tidyverse)
library(broom)
library(lfe)
library(parsnip)
```

### Step 1: Explore the Data

```{r data-exploration}
# Load data
data("gapminder")

# View structure and summary
glimpse(gapminder)
summary(gapminder)
```

### Step 2: Build a Basic Linear Model

```{r fit-model}
# Fit the linear model
lm_model <- linear_reg() %>%
  set_engine("lm") %>%
  fit(lifeExp ~ gdpPercap, data = gapminder)

# Print out tidy output
tidy(lm_model)
```

#### Interpretation
- **Association between GDP per capita and life expectancy:**  
  The coefficient for `gdpPercap` is **0.0007649**, indicating that for every 1-unit increase in GDP per capita, life expectancy increases by a very small amount (less than a day).  

- **Statistical significance:**  
  The relationship is statistically significant, as indicated by the small standard error.  


### Step 3: Transform GDP Per Capita (Log Transformation)

```{r log-transformation}
# Add a log-transformed GDP column
gapminder <- gapminder %>%
  mutate(log_gdpPercap = log(gdpPercap))

# Refit the model with log-transformed GDP
lm_log_model <- linear_reg() %>%
  set_engine("lm") %>%
  fit(lifeExp ~ log_gdpPercap, data = gapminder)

# Print out tidy output
tidy(lm_log_model)
```

#### Interpretation
- **Association between GDP per capita and life expectancy:**  
  After log-transforming GDP, the coefficient for `log_gdpPercap` is **8.4051**, showing that a 1% increase in GDP per capita is associated with an increase of **8.4 years** in life expectancy.  

- **Statistical significance:**  
  The relationship remains statistically significant, as the standard error is small relative to the coefficient.  

- **Change in the relationship compared to Step 2:**  
  The transformation changes the scale of the relationship. The effect size appears larger when expressed as a percentage increase, showing the diminishing returns of GDP growth on life expectancy.  


### Step 4: Visualize the Model Fit

```{r model-visualize}
# Scatter plot with regression line
gapminder %>%
  ggplot(aes(x = log_gdpPercap, y = lifeExp)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "blue") +
  labs(title = "Relationship Between Log GDP per Capita and Life Expectancy",
       x = "Log GDP per Capita",
       y = "Life Expectancy")
```


### Step 5: Add Fixed Effects for Continent

```{r fixed-effects}
# Fit the model with fixed effects
lm_fe_model <- linear_reg() %>%
  set_engine("lm") %>%
  fit(lifeExp ~ log_gdpPercap + continent, data = gapminder)

# Print out tidy output
tidy(lm_fe_model)

```

#### Interpretation
- **Association between GDP per capita and life expectancy:**  
  The coefficient for `log_gdpPercap` is **6.4220**, suggesting that a 1% increase in GDP per capita corresponds to an increase of **6.4 years** in life expectancy, accounting for continent-specific factors.  

- **Statistical significance:**  
  The relationship is still statistically significant.  

- **Change in the relationship compared to Step 3:**  
  The magnitude of the coefficient decreased, indicating that some variation in life expectancy attributed to GDP per capita is explained by continent-level factors.  

### Step 6: Use felm for country and continent Fixed Effects

```{r country-level-fixed}
# Fit the fixed effects model
felm_model <- felm(lifeExp ~ log_gdpPercap | country, data = gapminder)

# View the summary
tidy(felm_model)

```

#### Interpretation
- **Association between log GDP per capita and life expectancy:**  
  The coefficient for `log_gdpPercap` is **9.7690**, meaning a 1% increase in GDP per capita is associated with an increase of **9.8 years** in life expectancy after controlling for country-specific factors.  

- **Statistical significance:**  
  The relationship remains statistically significant, as indicated by the small standard error.  

- **Change in the relationship compared to Step 5:**  
  The coefficient increased compared to the continent fixed effects model, suggesting that variations within countries show a stronger relationship between GDP per capita and life expectancy. 
  
```{r global-trends}
felm_model_year <- felm(lifeExp ~ log_gdpPercap | country + year, data = gapminder)
tidy(felm_model_year)
```

- **Bonus:**  
  Adding year fixed effects would control for global trends, such as advancements in healthcare and international economic conditions, that influence both GDP and life expectancy over time.  
  
### Step 7: Compare Models

```{r compare-models}
# Tidy model outputs
tidy_lm <- tidy(lm_model)
tidy_lm_log <- tidy(lm_log_model)
tidy_lm_fe <- tidy(lm_fe_model)
tidy_felm <- tidy(felm_model)

# Combine for comparison
model_comparison <- bind_rows(
  tidy_lm %>% mutate(model = "Basic LM"),
  tidy_lm_log %>% mutate(model = "Log-transformed LM"),
  tidy_lm_fe %>% mutate(model = "LM with Continent FE"),
  tidy_felm %>% mutate(model = "FELM with Country FE")
)

# Filter for GDP term
model_comparison %>%
  filter(term == "log_gdpPercap" | term == "gdpPercap") %>%
  select(model, term, estimate, std.error, p.value)
```

#### Interpretation
- **Basic LM**: The initial linear model shows a very small positive association between GDP per capita and life expectancy.
- **Log-transformed LM**: Log-transforming GDP highlights a stronger relationship, reflecting diminishing returns.
- **LM with Continent FE**: Adding continent fixed effects adjusts for regional differences.
- **FELM with Country FE**: Introducing country-level fixed effects strengthens the association.
- **FELM with Country + Year FE**: Adding year fixed effects reduces the GDP coefficient, showing that time trends account for some variability.


### Step 8: Discuss and Interpret
#### Fixed Effects Impact
- Including fixed effects progressively refines the model:
  1. Controlling for continents reduces variability due to regional factors.  
  2. Country-specific effects isolate within-country trends.  
  3. Adding year fixed effects captures global time-related influences.  

#### Additional Factors to Improve the Model
1. **Healthcare:** Number of hospitals or healthcare spending.  
2. **Education:** Literacy rates or average schooling years.  
3. **Policy Stability:** Governance indicators.  
4. **Environmental Data:** Pollution and climate factors.  
5. **Technological Progress:** Internet penetration or innovation indices.  

Including these could enhance the explanatory power of the model and account for more variability in life expectancy.  
